### 2015-11-30 paper/1-introduction.tex
changes to paper

%? use paragraphs for layout instead of (dense) flowing text ?
%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
\subsection{DEVS}
%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms. Of interest in this paper are the 3 key formalisms: Classic \cite{ClassicDEVS}, Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation. This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a kernel's perspective only Atomic Models exist in the simulation linked to each other by connected ports.

@@

\subsection{Solution} %for lack of a better word
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing the adevs kernel hard.
The devs-ex-machina (dxex) kernel follows the design of the PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g. memory allocation strategy). 
The core aims of the project are to offer a deterministic simulation kernel where the simulation author is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalisms has $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. Dxex kernels can operate on IEEE754 floating point time units or integer time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
An integer representation significantly reduces the possible range of the virtual time, but avoids approximation errors. Furthermore, the notion of $\epsilon$ as a minimum distance between two distinct time points need not to be established. This task is non-trivial for floating point approximations.\\
The select() function, which imposes a sequential ordering between concurrent events, is made implicit in this project by extending time representation with a causality field with a range at least as large as the maximum nr of models in the kernels ($2^{24}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This avoids evaluating the select() function on all imminent models, while still maintaining the deterministic order of concurrent transitions.

>>>>

%? use paragraphs for layout instead of (dense) flowing text ?
%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
\subsection{DEVS}
%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms. Of interest in this paper are the 2 key formalisms: Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation. This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a simulation kernel's perspective only Atomic Models exist linked to each other by connected ports.

@@

\subsection{Solution} %for lack of a better word
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing the adevs kernel hard.
The devs-ex-machina (dxex) kernel follows the design of the PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g. memory allocation strategy). 
The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalisms has $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. Dxex kernels can operate on IEEE754 floating point time units or integer time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
An integer representation significantly reduces the possible range of the virtual time, but avoids approximation errors. Furthermore, the notion of $\epsilon$ as a minimum distance between two distinct time points need not to be established. This task is non-trivial for floating point approximations.\\
The select() function, which imposes a sequential ordering between concurrent events, is made implicit in this project by extending time representation with a causality field with a range at least as large as the maximum nr of models in the kernels ($2^{24}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This avoids evaluating the select() function on all imminent models, while still maintaining the deterministic order of concurrent transitions.

### 2015-11-30 paper/3-features.tex
changes to paper

%TODO explain that we are briefly going over the main features
\subsection{Based on PythonPDEVS}
%TODO say that it is based on PythonPDEVS
The simulator is based on PythonPDEVS, and provides the following features: 
\begin{enumerate}
	\item Direct Connection
	\item Dynamic Structured DEVS
	\item Termination function. If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate. Only available in single-threaded simulation.
	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
	\item Optimistic and Conservative synchronization of PDEVS.
\end{enumerate}
The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use. For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.\\
The tracing is not comparable to adevs's listener interface. To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points. Furthermore, the framework itself has to be thread safe and deterministic so that a simulation will always produce the exact same output.
%TODO Decide whether we should list the functionality that is _not_ implemented
The following features from PyPDEVS are not present
\begin{enumerate}
	\item Activity tracking and relocation
	\item Serialization
	\item Interactive control. 
	\item Distributed simulation
\end{enumerate}
Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation. If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
While PyPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.

>>>>

%TODO explain that we are briefly going over the main features
\subsection{Based on PythonPDEVS}
%TODO say that it is based on PythonPDEVS
The simulator is based on PythonPDEVS, and provides the following features: 
\begin{enumerate}
	\item Direct Connection
	\item Dynamic Structured DEVS
	\item Termination function. If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate. Only available in single-threaded simulation.
	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
	\item Optimistic and Conservative synchronization of PDEVS.
\end{enumerate}
The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use. For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use. Another example is the state saving mechanism, which is only used for optimistic parallel simulation and has no performance impact in conservative parallel simulation.\\
The tracing is not comparable to adevs's listener interface. To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points. Furthermore, the framework itself has to be thread safe and deterministic so that a simulation will always produce the exact same output.
%TODO Decide whether we should list the functionality that is _not_ implemented
The following features from PyPDEVS are not present
\begin{enumerate}
	\item Activity tracking and relocation
	\item Serialization
	\item Interactive control. 
	\item Distributed simulation
\end{enumerate}
Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp. %This state saving has no impact in a single threaded or conservative kernel.
\\
Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation. If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.
